---
title: "Objective_3_RM_analysis"
format: html
editor: visual
---

## Setup

Adjust `input_path` below to point at your combined RM annotation CSV.

```{r}
#| label: setup

library(tidyverse)
library(janitor)
library(gt)

input_path <- "~/Objective_3/Objective_3_RM_analysis/data/all.rm.genes.concatRpt.tsv"   # <-- change this to your actual file path

df <- read_tsv(input_path, show_col_types = FALSE) %>%
  clean_names()   # -> sample, operon, gene, system_type, gene_type,
                  #    hmm, evalue, predicted_methylation,
                  #    prediction_confidence, rebase_homolog,
                  #    homolog_identity_percent, homolog_methylation,
                  #    homolog_motif

# remove "fix_" prefix from sample IDs (e.g. "fix_EcCa437a" -> "EcCa437a")
# so they match the metadata's Isolate_ID directly
df <- df %>%
  mutate(sample = str_remove(sample, "^fix_"))

# derive contig from the Gene locus tag, e.g. "contig_1_2542" -> "contig_1"
df <- df %>%
  mutate(contig = str_extract(gene, "^contig_[0-9]+"))

n_isolates <- n_distinct(df$sample)
n_hmm <- n_distinct(df$hmm)
```

This dataset covers **`r n_isolates`** isolates and **`r n_hmm`** distinct HMM models.

## 1. Gene type counts per isolate, per contig

Counts of each gene type (RE, MT, SP, IIG) found on each contig, within each isolate.

```{r}
#| label: gene-type-by-contig

gene_type_by_contig <- df %>%
  count(sample, contig, gene_type) %>%
  pivot_wider(names_from = gene_type, values_from = n, values_fill = 0) %>%
  arrange(sample, contig)

gene_type_by_contig
```

## 1b. Chromosome vs plasmid tables

`contig_1` is treated as the chromosome; every other contig is treated as a plasmid. Isolates can have multiple plasmid contigs, so the plasmid table aggregates gene type counts per isolate across all non-chromosome contigs (with a `n_plasmid_contigs` column showing how many distinct plasmid contigs contributed).

```{r}
#| label: chromosome-vs-plasmid

df <- df %>%
  mutate(replicon_type = if_else(contig == "contig_1", "Chromosome", "Plasmid"))

# Chromosome table: one row per isolate, gene type counts on contig_1 only
chromosome_gene_types <- df %>%
  filter(replicon_type == "Chromosome") %>%
  count(sample, gene_type) %>%
  pivot_wider(names_from = gene_type, values_from = n, values_fill = 0) %>%
  arrange(sample)

# Plasmid table: one row per isolate, gene type counts summed across all
# non-chromosome contigs, plus a count of how many plasmid contigs were found
plasmid_gene_types <- df %>%
  filter(replicon_type == "Plasmid") %>%
  group_by(sample) %>%
  mutate(n_plasmid_contigs = n_distinct(contig)) %>%
  ungroup() %>%
  count(sample, n_plasmid_contigs, gene_type) %>%
  pivot_wider(names_from = gene_type, values_from = n, values_fill = 0) %>%
  arrange(sample)

chromosome_gene_types
plasmid_gene_types
```

## 2. System type totals per isolate

Counts of each system type (RM_Type_I, RM_Type_II, RM_Type_IIG, Type_IV) per isolate, with a row total.

```{r}
#| label: system-type-totals

system_type_totals <- df %>%
  count(sample, system_type) %>%
  pivot_wider(names_from = system_type, values_from = n, values_fill = 0) %>%
  mutate(Total = rowSums(across(where(is.numeric)))) %>%
  arrange(sample)

system_type_totals
```

## 3. Operon-level summary

One row per isolate x operon, showing the operon's system type(s), contig(s), gene types present, gene count, and whether any gene in the operon has a methylation call.

```{r}
#| label: operon-summary

operon_summary <- df %>%
  group_by(sample, operon) %>%
  summarise(
    contigs        = paste(sort(unique(contig)), collapse = ";"),
    operon_type    = paste(sort(unique(system_type)), collapse = ";"),
    gene_types     = paste(sort(unique(gene_type)), collapse = ";"),
    n_genes        = n(),
    has_methylation_call = any(!is.na(predicted_methylation) & predicted_methylation != ""),
    .groups = "drop"
  ) %>%
  arrange(sample, operon)

operon_summary
```

## 4. HMM inventory

### 4a. Overall counts across all isolates

```{r}
#| label: hmm-overall

hmm_inventory_overall <- df %>%
  count(hmm, sort = TRUE) %>%
  rename(n_occurrences = n)

hmm_inventory_overall %>%
  gt() %>%
  tab_header(title = "Distinct HMM models identified across all isolates")
```

### 4b. Per-isolate breakdown

```{r}
#| label: hmm-by-isolate

hmm_inventory_by_isolate <- df %>%
  count(sample, hmm) %>%
  pivot_wider(names_from = hmm, values_from = n, values_fill = 0) %>%
  arrange(sample)

hmm_inventory_by_isolate
```

::: callout-note
The raw `hmm` column already condenses to \~16 distinct models across the whole dataset — no further grouping needed. Note that HMM family suffixes (e.g. `_FAM_4` vs `_FAM_1`) denote genuinely different, unrelated HMM families rather than variants of the same model, so they should **not** be collapsed together.
:::

## 5. Heatmaps

### 5a. HMM counts per isolate — ggplot2

A tile heatmap of every isolate x HMM combination, coloured by how many copies of that HMM were found. Missing combinations are filled with 0 (absence) via `complete()`, so every isolate x HMM cell is represented even if it never appeared in the raw data.

```{r}
#| label: hmm-heatmap-ggplot
#| fig-width: 10
#| fig-height: 18

hmm_counts_long <- df %>%
  count(sample, hmm) %>%
  complete(sample, hmm, fill = list(n = 0))

ggplot(hmm_counts_long, aes(x = hmm, y = sample, fill = n)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_gradient(low = "grey95", high = "darkred", name = "Copies") +
  labs(x = "HMM model", y = "Isolate",
       title = "HMM copy number per isolate") +
  theme_minimal(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 4),   # 139 isolates - keep labels tiny
    panel.grid = element_blank()
  )
```

With 139 isolates the y-axis labels will be very small — this view is mainly for spotting overall patterns/outliers rather than reading individual isolate names off the plot. Use the `pheatmap` version below to reorder isolates by similarity, which usually makes patterns much easier to see, or filter `hmm_counts_long` to a subset of isolates first if you want a readable static figure.

### 5b. HMM presence/absence with clustering — pheatmap

`pheatmap` automatically performs hierarchical clustering on both rows (isolates) and columns (HMMs), so isolates with similar RM system profiles group together, and reveals which HMMs tend to co-occur. This uses presence/absence (0/1) rather than raw counts, which is usually cleaner for clustering — switch `binary_matrix` for `count_matrix` (commented below) if you want clustering based on copy number instead.

```{r}
#| label: hmm-heatmap-pheatmap
#| fig-width: 10
#| fig-height: 20

library(pheatmap)

# Wide matrix: isolates as rows, HMMs as columns
hmm_wide <- df %>%
  count(sample, hmm) %>%
  pivot_wider(names_from = hmm, values_from = n, values_fill = 0)

count_matrix <- hmm_wide %>%
  select(-sample) %>%
  as.matrix()
rownames(count_matrix) <- hmm_wide$sample

binary_matrix <- (count_matrix > 0) * 1   # presence/absence

pheatmap(
  binary_matrix,                 # swap for count_matrix to cluster on copy number
  color = c("grey95", "darkred"),
  legend_breaks = c(0, 1),
  legend_labels = c("Absent", "Present"),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize_row = 3,               # small - 139 isolates
  fontsize_col = 8,
  main = "HMM presence/absence across isolates (clustered)"
)
```

If `pheatmap` isn't installed, run `install.packages("pheatmap")` first.

## 6. HMM heatmap with metadata colour strips

This adds isolate-level metadata (e.g. species, source, collection site) as colour strips alongside the HMM heatmap from 5a. Uses `patchwork` to stack the strips and the heatmap so isolate rows line up exactly.

**Adjust the placeholders marked below to match your actual metadata file and column names.**

```{r}
#| label: hmm-heatmap-metadata
#| fig-width: 11
#| fig-height: 18

library(patchwork)

# ---- 1. Read metadata -----------------------------------------------
metadata_path <- "data/isolate_metadata.csv"   # <-- change to your actual metadata file path
metadata <- read_csv("~/Objective_3/Objective_3_RM_analysis/data/metadata_Apr26.csv", show_col_types = FALSE) %>%
  clean_names()   # -> isolate_id, farm_id, sample_site, total_length, total_contigs,
                   #    gc_percent, n50, l90, sbi, lactose, sorbitol, surcrose, stx_gene

# Join key: df$sample and metadata's isolate_id now match directly
# (both e.g. "EcCa437a") since the "fix_" prefix was stripped in Setup.
metadata <- metadata %>%
  rename(sample = isolate_id) %>%
  mutate(farm_id = as.factor(farm_id))   # farm ID is categorical, not numeric

# Columns to show as colour strips - add/remove as needed
annotation_vars <- c("sample_site", "sbi", "stx_gene")

# ---- 2. Join metadata onto the HMM long-format counts ----------------
hmm_counts_long_annot <- df %>%
  count(sample, hmm) %>%
  complete(sample, hmm, fill = list(n = 0)) %>%
  left_join(metadata, by = "sample")

# Fix isolate ordering so every panel below shares the same y-axis order
isolate_order <- sort(unique(hmm_counts_long_annot$sample))
hmm_counts_long_annot <- hmm_counts_long_annot %>%
  mutate(sample = factor(sample, levels = isolate_order))

# ---- 3. Main HMM heatmap ---------------------------------------------
main_heatmap <- ggplot(hmm_counts_long_annot, aes(x = hmm, y = sample, fill = n)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_gradient(low = "grey95", high = "darkred", name = "Copies") +
  labs(x = "HMM model", y = NULL) +
  theme_minimal(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 4),
    panel.grid = element_blank(),
    legend.position = "right"
  )

# ---- 4. One narrow colour-strip panel per annotation variable ---------
make_strip <- function(var) {
  strip_data <- hmm_counts_long_annot %>%
    distinct(sample, .data[[var]])

  ggplot(strip_data, aes(x = 1, y = sample, fill = .data[[var]])) +
    geom_tile(color = "white", linewidth = 0.1) +
    labs(x = NULL, y = NULL, fill = var) +
    theme_minimal(base_size = 9) +
    theme(
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      legend.position = "right"
    )
}

strip_plots <- lapply(annotation_vars, make_strip)

# ---- 5. Combine strips + main heatmap side by side --------------------
# Strips are narrow, main heatmap is wide - widths reflect that
combined_plot <- wrap_plots(
  c(strip_plots, list(main_heatmap)),
  nrow = 1,
  widths = c(rep(0.3, length(strip_plots)), 10)
)

combined_plot
```

If `patchwork` isn't installed, run `install.packages("patchwork")` first. The `make_strip()` helper works for any number of annotation variables — just add or remove names from `annotation_vars`, no other code changes needed.

## 7. System type heatmap

Same treatment as the HMM heatmap in Section 5, but using the `system_type_totals` table (RM_Type_I, RM_Type_II, RM_Type_IIG, Type_IV counts per isolate) instead of individual HMMs. With only 4-5 system type columns this is a much narrower heatmap than the HMM one.

### 7a. System type counts per isolate — ggplot2

```{r}
#| label: system-type-heatmap-ggplot
#| fig-width: 6
#| fig-height: 18

system_type_long <- system_type_totals %>%
  select(-Total) %>%
  pivot_longer(-sample, names_to = "system_type", values_to = "n")

ggplot(system_type_long, aes(x = system_type, y = sample, fill = n)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_gradient(low = "grey95", high = "darkblue", name = "Count") +
  labs(x = "System type", y = "Isolate",
       title = "RM system type counts per isolate") +
  theme_minimal(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 4),   # 139 isolates - keep labels tiny
    panel.grid = element_blank()
  )
```

### 7b. System type counts with clustering — pheatmap

```{r}
#| label: system-type-heatmap-pheatmap
#| fig-width: 8
#| fig-height: 20

system_type_matrix <- system_type_totals %>%
  select(-Total) %>%
  column_to_rownames("sample") %>%
  as.matrix()

pheatmap(
  system_type_matrix,
  color = colorRampPalette(c("grey95", "darkblue"))(50),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize_row = 3,               # small - 139 isolates
  fontsize_col = 9,
  main = "RM system type counts across isolates (clustered)"
)
```

## 8. Gene type heatmap (by isolate, not contig)

Same idea as Section 1's `gene_type_by_contig`, but collapsed across contigs so there's one row per isolate rather than one row per isolate x contig. This gives total counts of each gene type (RE, MT, SP, IIG) per isolate, regardless of which contig they're on.

```{r}
#| label: gene-type-by-isolate

gene_type_by_isolate <- df %>%
  count(sample, gene_type) %>%
  pivot_wider(names_from = gene_type, values_from = n, values_fill = 0) %>%
  arrange(sample)

gene_type_by_isolate
```

### 8a. Gene type counts per isolate — ggplot2

```{r}
#| label: gene-type-heatmap-ggplot
#| fig-width: 6
#| fig-height: 18

gene_type_long <- gene_type_by_isolate %>%
  pivot_longer(-sample, names_to = "gene_type", values_to = "n")

ggplot(gene_type_long, aes(x = gene_type, y = sample, fill = n)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_gradient(low = "grey95", high = "darkgreen", name = "Count") +
  labs(x = "Gene type", y = "Isolate",
       title = "Gene type counts per isolate") +
  theme_minimal(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 4),   # 139 isolates - keep labels tiny
    panel.grid = element_blank()
  )
```

### 8b. Gene type counts with clustering — pheatmap

```{r}
#| label: gene-type-heatmap-pheatmap
#| fig-width: 8
#| fig-height: 20

gene_type_matrix <- gene_type_by_isolate %>%
  column_to_rownames("sample") %>%
  as.matrix()

pheatmap(
  gene_type_matrix,
  color = colorRampPalette(c("grey95", "darkgreen"))(50),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize_row = 3,               # small - 139 isolates
  fontsize_col = 9,
  main = "Gene type counts across isolates (clustered)"
)
```

## 9. PCoA of HMM profiles

Rather than a strict 0/1 presence/absence matrix, this builds an isolate x HMM **count** matrix (copy number of each HMM per isolate), then uses **Bray-Curtis dissimilarity** so isolates that differ in *how many* copies of an HMM they carry (not just whether they carry it at all) end up further apart in the ordination.

```{r}
#| label: hmm-dist-matrix
#| fig-width: 8
#| fig-height: 7

library(vegan)
library(ape)

# ---- 1. Build isolate x HMM count matrix -------------------------
hmm_pcoa_wide <- df %>%
  count(sample, hmm) %>%
  pivot_wider(names_from = hmm, values_from = n, values_fill = 0)

hmm_pcoa_matrix <- hmm_pcoa_wide %>%
  select(-sample) %>%
  as.matrix()
rownames(hmm_pcoa_matrix) <- hmm_pcoa_wide$sample

# ---- 2. Bray-Curtis dissimilarity (weights by copy number) -----
hmm_dist <- vegdist(hmm_pcoa_matrix, method = "bray")

# View as a regular isolate x isolate matrix
hmm_dist_matrix <- as.matrix(hmm_dist)
hmm_dist_matrix[1:5, 1:5]   # preview - full matrix is 139 x 139
```

### 9a. Distance matrix heatmap

A quick visual check of the distance matrix itself before running the PCoA — darker cells mean more dissimilar HMM profiles between that pair of isolates. Diagonal is always 0 (an isolate compared to itself).

```{r}
#| label: hmm-dist-heatmap
#| fig-width: 10
#| fig-height: 10

pheatmap(
  hmm_dist_matrix,
  color = colorRampPalette(c("white", "firebrick"))(50),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize_row = 3,
  fontsize_col = 3,
  main = "Bray-Curtis dissimilarity between isolates (HMM copy number)"
)
```

### 9b. PCoA

```{r}
#| label: hmm-pcoa
#| fig-width: 8
#| fig-height: 7

# ---- 3. PCoA -------------------------------------------------------
# Bray-Curtis is non-Euclidean, so classical PCoA can produce negative
# eigenvalues - when that happens, the raw Relative_eig percentages can
# sum to over 100% (a known artefact, not an error). The Cailliez
# correction shifts all eigenvalues non-negative first, so the corrected
# percentages (Rel_corr_eig) behave properly and sum to <=100%.
hmm_pcoa <- pcoa(hmm_dist, correction = "cailliez")

# Check whether a correction was actually needed/applied
hmm_pcoa$note

# % variance explained by first two axes - use corrected values if present
if ("Rel_corr_eig" %in% colnames(hmm_pcoa$values)) {
  var_explained <- round(hmm_pcoa$values$Rel_corr_eig[1:2] * 100, 1)
} else {
  var_explained <- round(hmm_pcoa$values$Relative_eig[1:2] * 100, 1)
}

pcoa_scores <- as.data.frame(hmm_pcoa$vectors[, 1:2])
colnames(pcoa_scores) <- c("PCo1", "PCo2")
pcoa_scores$sample <- rownames(pcoa_scores)

# Join sample_site from metadata (built in Section 6) so points can be coloured
pcoa_scores <- pcoa_scores %>%
  left_join(metadata %>% select(sample, sample_site), by = "sample")

# ---- 4. Add jittered coordinates (fixed seed) ------------------------
# Baked into pcoa_scores itself, not applied via ggplot's position_jitter,
# so any later plot reusing PCo1_jitter/PCo2_jitter keeps points in the
# exact same place as this plot.
set.seed(42)
jitter_amount_x <- diff(range(pcoa_scores$PCo1)) * 0.05
jitter_amount_y <- diff(range(pcoa_scores$PCo2)) * 0.05

pcoa_scores <- pcoa_scores %>%
  mutate(
    PCo1_jitter = jitter(PCo1, amount = jitter_amount_x),
    PCo2_jitter = jitter(PCo2, amount = jitter_amount_y)
  )

# ---- 5. Custom colour palette ---------------------------------------
Sample_site_colours <- c(
  "Farm"     = "#88ccee",
  "Lairage" = "#44aa99",
  "Hide"  = "#DDCC77",
  "Pre_intervention" = "#cc6677",
  "Post_intervention" = "#332288"
)

# ---- 6. Plot -----------------------------------------------------------
ggplot(pcoa_scores, aes(x = PCo1_jitter, y = PCo2_jitter, color = sample_site)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = Sample_site_colours, name = "Sample site") +
  labs(
    x = paste0("PCo1 (", var_explained[1], "%, jittered)"),
    y = paste0("PCo2 (", var_explained[2], "%, jittered)"),
    title = "PCoA of HMM copy-number profiles (Bray-Curtis dissimilarity)"
  ) +
  theme_minimal(base_size = 11)
```


If `vegan` or `ape` aren't installed, run `install.packages(c("vegan", "ape"))` first.

Points that sit on top of each other or very close together represent isolates with near-identical HMM profiles (e.g. your `a`/`b`/`c` replicate series are a good candidate to check for tight clustering here). Two axes are plotted by default (`PCo1`/`PCo2`), but `hmm_pcoa$values` holds the full eigenvalue table if you want to check how much variance further axes explain, and `pcoa_scores` can be extended to plot PCo1 vs PCo3, etc.

Points are jittered (small random offset) so overlapping isolates are still visible as separate dots, and coloured by `sample_site` using your custom palette (`Sample_site_colours`). The jitter uses a fixed seed (`set.seed(42)`) and is stored directly in `pcoa_scores$PCo1_jitter`/`PCo2_jitter`, so any later plot you build off `pcoa_scores` in this document will place the same isolate at the same position every time — just reuse those two columns rather than re-jittering `PCo1`/`PCo2` freshly, which would give a different random offset. To colour by a different metadata variable instead (e.g. `sbi`, `farm_id`, `stx_gene`), swap `sample_site` for that column name in both the `select()` in the join and the `aes(color = ...)` call, and provide a matching named colour vector for it.

```{r}
#| label: hmm-pcoa-sbi
#| fig-width: 8
#| fig-height: 7

# Join sbi from metadata onto the existing pcoa_scores
pcoa_scores <- pcoa_scores %>%
  left_join(metadata %>% select(sample, sbi), by = "sample")

# ---- Custom colour palette ---------------------------------------
SBI_colours <- c(
  "AWY2a" = "#2CA02C",
  "AY2a"  = "#1F77B4",
  "WY12a" = "#AEC7E8",
  "ASY2c" = "#FF7F0E",
  "SY2c"  = "#FFBB78"
)

# ---- Plot -----------------------------------------------------------
ggplot(pcoa_scores, aes(x = PCo1_jitter, y = PCo2_jitter, color = sbi)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = SBI_colours, name = "SBI") +
  labs(
    x = paste0("PCo1 (", var_explained[1], "%, jittered)"),
    y = paste0("PCo2 (", var_explained[2], "%, jittered)"),
    title = "PCoA of HMM copy-number profiles (Bray-Curtis dissimilarity)"
  ) +
  theme_minimal(base_size = 11)
```

### 9d. PCoA coloured by Stx gene

Same underlying points again (identical `PCo1_jitter`/`PCo2_jitter` positions), coloured by `stx_gene` instead.

```{r}
#| label: hmm-pcoa-stx
#| fig-width: 8
#| fig-height: 7

# Join stx_gene from metadata onto the existing pcoa_scores
pcoa_scores <- pcoa_scores %>%
  left_join(metadata %>% select(sample, stx_gene), by = "sample")

# ---- Custom colour palette ---------------------------------------
Stx_colours <- c(
  "Absent" = "#FF0800",
  "Stx2c" = "#6610F2",
  "Stx1a_and_2a" = "#410200",
  "Stx2a"  = "#F17105"
)

# ---- Plot -----------------------------------------------------------
ggplot(pcoa_scores, aes(x = PCo1_jitter, y = PCo2_jitter, color = stx_gene)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = Stx_colours, name = "Stx gene") +
  labs(
    x = paste0("PCo1 (", var_explained[1], "%, jittered)"),
    y = paste0("PCo2 (", var_explained[2], "%, jittered)"),
    title = "PCoA of HMM copy-number profiles (Bray-Curtis dissimilarity)"
  ) +
  theme_minimal(base_size = 11)
```


## 10. SBI-level condensed summary (Table 1 style)

Modelled on Table 1 of [Galeone et al. 2025](https://pmc.ncbi.nlm.nih.gov/articles/PMC12016153/), which lists the actual detected motifs per sample rather than counts. Here, one row per `SBI` type, with the actual distinct gene types, system types, and HMMs found across all isolates of that SBI written out (comma-separated) rather than tallied.

```{r}
#| label: sbi-condensed-table

sbi_condensed <- df %>%
  left_join(metadata %>% select(sample, sbi), by = "sample") %>%
  group_by(sbi) %>%
  summarise(
    `Gene type`   = paste(sort(unique(gene_type)), collapse = ", "),
    `System type` = paste(sort(unique(system_type)), collapse = ", "),
    HMM           = paste(sort(unique(hmm)), collapse = ", "),
    .groups = "drop"
  ) %>%
  rename(SBI = sbi) %>%
  arrange(SBI)

sbi_condensed %>%
  gt() %>%
  tab_header(title = "RM system components by SBI type") %>%
  cols_align(align = "left", columns = everything())
```


## 10c. SBI-level condensed summary with % of isolates

Since SBI groups contain very different numbers of isolates, raw counts (10b) aren't directly comparable across groups. This version instead shows, for each value, **the percentage of isolates within that SBI group that carry it at least once** (presence-based, not summed occurrences) — a fairer basis for comparing SBI types of different sizes.

```{r}
#| label: sbi-condensed-table-pct

# Rebuilt here so this chunk doesn't depend on 10b having run first
sbi_meta <- df %>%
  left_join(metadata %>% select(sample, sbi), by = "sample")

# Number of isolates per SBI (restricted to isolates actually present in the RM data)
sbi_n_isolates <- metadata %>%
  filter(sample %in% unique(df$sample)) %>%
  distinct(sample, sbi) %>%
  count(sbi, name = "n_isolates")

# Helper: % of isolates in each SBI group carrying each distinct value at least once
pct_isolates_with <- function(value_col, out_name) {
  sbi_meta %>%
    distinct(sample, sbi, {{ value_col }}) %>%
    count(sbi, {{ value_col }}, name = "n_with") %>%
    left_join(sbi_n_isolates, by = "sbi") %>%
    mutate(pct = round(100 * n_with / n_isolates, 0)) %>%
    arrange(sbi, {{ value_col }}) %>%
    group_by(sbi) %>%
    summarise(
      "{out_name}" := paste0({{ value_col }}, " (", pct, "%)", collapse = ", "),
      .groups = "drop"
    )
}

sbi_gene_type_pct   <- pct_isolates_with(gene_type, "Gene type")
sbi_system_type_pct <- pct_isolates_with(system_type, "System type")
sbi_hmm_pct         <- pct_isolates_with(hmm, "HMM")

sbi_condensed_pct <- sbi_gene_type_pct %>%
  left_join(sbi_system_type_pct, by = "sbi") %>%
  left_join(sbi_hmm_pct, by = "sbi") %>%
  left_join(sbi_n_isolates, by = "sbi") %>%
  rename(SBI = sbi, `n isolates` = n_isolates) %>%
  arrange(SBI)

sbi_condensed_pct %>%
  gt() %>%
  tab_header(title = "RM system components by SBI type (% of isolates)") %>%
  cols_align(align = "left", columns = everything())
```



## 10d. SBI-level condensed summary with mean occurrences per isolate

Rather than presence (10c), this shows the **average number of occurrences per isolate** within each SBI group — e.g. if `GmrSD_N` averages 2.4 in a group, isolates in that group typically carry more than one copy. Isolates with zero copies of a value are included as 0 in the average (via `complete()`), not skipped, so this is a true per-isolate mean across the whole group, not just among isolates that happen to carry it.

```{r}
#| label: sbi-condensed-table-mean

# Helper: mean occurrences per isolate within each SBI group
mean_occurrences <- function(value_col, out_name) {
  per_isolate_counts <- sbi_meta %>%
    count(sbi, sample, {{ value_col }}) %>%
    complete(nesting(sbi, sample), {{ value_col }}, fill = list(n = 0))

  per_isolate_counts %>%
    group_by(sbi, {{ value_col }}) %>%
    summarise(mean_n = round(mean(n), 1), .groups = "drop") %>%
    arrange(sbi, {{ value_col }}) %>%
    group_by(sbi) %>%
    summarise(
      "{out_name}" := paste0({{ value_col }}, " (", mean_n, ")", collapse = ", "),
      .groups = "drop"
    )
}

sbi_gene_type_mean   <- mean_occurrences(gene_type, "Gene type")
sbi_system_type_mean <- mean_occurrences(system_type, "System type")
sbi_hmm_mean         <- mean_occurrences(hmm, "HMM")

sbi_condensed_mean <- sbi_gene_type_mean %>%
  left_join(sbi_system_type_mean, by = "sbi") %>%
  left_join(sbi_hmm_mean, by = "sbi") %>%
  left_join(sbi_n_isolates, by = "sbi") %>%
  rename(SBI = sbi, `n isolates` = n_isolates) %>%
  arrange(SBI)

sbi_condensed_mean %>%
  gt() %>%
  tab_header(title = "RM system components by SBI type (mean occurrences per isolate)") %>%
  cols_align(align = "left", columns = everything())

write_csv(sbi_condensed_mean, "data/sbi_condensed_mean.csv")
```


## 10e. Long-format comparison table (one row per SBI, one column per value)

Each SBI type gets one row, with a separate column for every `Gene type`/`System type`/`HMM` value holding its mean occurrence per isolate, plus an `n isolates` column showing how many isolates that row is based on.

```{r}
#| label: sbi-long-table

# Helper: mean occurrences per isolate, long format (one row per sbi x value)
build_category_means_long <- function(value_col, category_name) {
  sbi_meta %>%
    count(sbi, sample, {{ value_col }}) %>%
    complete(nesting(sbi, sample), {{ value_col }}, fill = list(n = 0)) %>%
    group_by(sbi, {{ value_col }}) %>%
    summarise(mean_n = round(mean(n), 1), .groups = "drop") %>%
    rename(Value = {{ value_col }}) %>%
    mutate(Category = category_name, .before = 1)
}

gene_type_long   <- build_category_means_long(gene_type, "Gene type")
system_type_long <- build_category_means_long(system_type, "System type")
hmm_long         <- build_category_means_long(hmm, "HMM")

# Order rows by category first (Gene type, System type, HMM), so columns
# appear grouped in that order once pivoted wide below
sbi_long_all <- bind_rows(gene_type_long, system_type_long, hmm_long) %>%
  mutate(Category = factor(Category, levels = c("Gene type", "System type", "HMM"))) %>%
  arrange(Category, Value)

# Transpose: one row per SBI, one column per value
sbi_wide_table <- sbi_long_all %>%
  select(sbi, Value, mean_n) %>%
  pivot_wider(names_from = Value, values_from = mean_n, values_fill = 0) %>%
  left_join(sbi_n_isolates, by = "sbi") %>%
  rename(SBI = sbi, `n isolates` = n_isolates) %>%
  relocate(`n isolates`, .after = SBI) %>%
  arrange(SBI)

# Column names belonging to each category, for the spanner labels below
gene_type_cols   <- unique(gene_type_long$Value)
system_type_cols <- unique(system_type_long$Value)
hmm_cols         <- unique(hmm_long$Value)

sbi_wide_table %>%
  gt() %>%
  tab_header(title = "Mean occurrences per isolate by SBI type") %>%
  tab_spanner(label = "Gene type", columns = all_of(gene_type_cols)) %>%
  tab_spanner(label = "System type", columns = all_of(system_type_cols)) %>%
  tab_spanner(label = "HMM", columns = all_of(hmm_cols)) %>%
  cols_align(align = "left", columns = SBI)

write_csv(sbi_wide_table, "data/sbi_wide_table.csv")
```


## 10f. Heatmap version

Same data as 10e's table, as a heatmap instead. `facet_grid()` on `Category` reproduces the Gene type/System type/HMM grouping as strip labels above each block of columns, matching the spanners in the table. Value labels are shown directly in each tile since there are only 5 SBI rows.

```{r}
#| label: sbi-heatmap
#| fig-width: 14
#| fig-height: 4

ggplot(sbi_long_all, aes(x = Value, y = sbi, fill = mean_n)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = mean_n), size = 2.8, color = "black") +
  scale_fill_gradient(low = "grey95", high = "darkblue", name = "Mean\noccurrences") +
  facet_grid(~ Category, scales = "free_x", space = "free_x") +
  labs(x = NULL, y = "SBI") +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "grey90", color = NA),
    strip.text = element_text(face = "bold"),
    panel.grid = element_blank()
  )
```