---
title: "Objective_3_RM_analysis"
format: html
editor: visual
---

## Setup

Adjust `input_path` below to point at your combined RM annotation CSV.

```{r}
#| label: setup

library(tidyverse)
library(janitor)
library(gt)

input_path <- "~/Objective_3/Objective_3_RM_analysis/data/all.rm.genes.concatRpt.tsv"   # <-- change this to your actual file path

df <- read_tsv(input_path, show_col_types = FALSE) %>%
  clean_names()   # -> sample, operon, gene, system_type, gene_type,
                  #    hmm, evalue, predicted_methylation,
                  #    prediction_confidence, rebase_homolog,
                  #    homolog_identity_percent, homolog_methylation,
                  #    homolog_motif

# remove "fix_" prefix from sample IDs (e.g. "fix_EcCa437a" -> "EcCa437a")
# so they match the metadata's Isolate_ID directly
df <- df %>%
  mutate(sample = str_remove(sample, "^fix_"))

# derive contig from the Gene locus tag, e.g. "contig_1_2542" -> "contig_1"
df <- df %>%
  mutate(contig = str_extract(gene, "^contig_[0-9]+"))

n_isolates <- n_distinct(df$sample)
n_hmm <- n_distinct(df$hmm)
```

This dataset covers **`r n_isolates`** isolates and **`r n_hmm`** distinct HMM models.

## 1. Gene type counts per isolate, per contig

Counts of each gene type (RE, MT, SP, IIG) found on each contig, within each isolate.

```{r}
#| label: gene-type-by-contig

gene_type_by_contig <- df %>%
  count(sample, contig, gene_type) %>%
  pivot_wider(names_from = gene_type, values_from = n, values_fill = 0) %>%
  arrange(sample, contig)

gene_type_by_contig
```

## 1b. Chromosome vs plasmid tables

`contig_1` is treated as the chromosome; every other contig is treated as a plasmid. Isolates can have multiple plasmid contigs, so the plasmid table aggregates gene type counts per isolate across all non-chromosome contigs (with a `n_plasmid_contigs` column showing how many distinct plasmid contigs contributed).

```{r}
#| label: chromosome-vs-plasmid

df <- df %>%
  mutate(replicon_type = if_else(contig == "contig_1", "Chromosome", "Plasmid"))

# Chromosome table: one row per isolate, gene type counts on contig_1 only
chromosome_gene_types <- df %>%
  filter(replicon_type == "Chromosome") %>%
  count(sample, gene_type) %>%
  pivot_wider(names_from = gene_type, values_from = n, values_fill = 0) %>%
  arrange(sample)

# Plasmid table: one row per isolate, gene type counts summed across all
# non-chromosome contigs, plus a count of how many plasmid contigs were found
plasmid_gene_types <- df %>%
  filter(replicon_type == "Plasmid") %>%
  group_by(sample) %>%
  mutate(n_plasmid_contigs = n_distinct(contig)) %>%
  ungroup() %>%
  count(sample, n_plasmid_contigs, gene_type) %>%
  pivot_wider(names_from = gene_type, values_from = n, values_fill = 0) %>%
  arrange(sample)

chromosome_gene_types
plasmid_gene_types
```

## 2. System type totals per isolate

Counts of each system type (RM_Type_I, RM_Type_II, RM_Type_IIG, Type_IV) per isolate, with a row total.

```{r}
#| label: system-type-totals

system_type_totals <- df %>%
  count(sample, system_type) %>%
  pivot_wider(names_from = system_type, values_from = n, values_fill = 0) %>%
  mutate(Total = rowSums(across(where(is.numeric)))) %>%
  arrange(sample)

system_type_totals
```

## 3. Operon-level summary

One row per isolate x operon, showing the operon's system type(s), contig(s), gene types present, gene count, and whether any gene in the operon has a methylation call.

```{r}
#| label: operon-summary

operon_summary <- df %>%
  group_by(sample, operon) %>%
  summarise(
    contigs        = paste(sort(unique(contig)), collapse = ";"),
    operon_type    = paste(sort(unique(system_type)), collapse = ";"),
    gene_types     = paste(sort(unique(gene_type)), collapse = ";"),
    n_genes        = n(),
    has_methylation_call = any(!is.na(predicted_methylation) & predicted_methylation != ""),
    .groups = "drop"
  ) %>%
  arrange(sample, operon)

operon_summary
```

## 4. HMM inventory

### 4a. Overall counts across all isolates

```{r}
#| label: hmm-overall

hmm_inventory_overall <- df %>%
  count(hmm, sort = TRUE) %>%
  rename(n_occurrences = n)

hmm_inventory_overall %>%
  gt() %>%
  tab_header(title = "Distinct HMM models identified across all isolates")
```

### 4b. Per-isolate breakdown

```{r}
#| label: hmm-by-isolate

hmm_inventory_by_isolate <- df %>%
  count(sample, hmm) %>%
  pivot_wider(names_from = hmm, values_from = n, values_fill = 0) %>%
  arrange(sample)

hmm_inventory_by_isolate
```

::: {.callout-note}
The raw `hmm` column already condenses to ~16 distinct models across the whole dataset — no further grouping needed. Note that HMM family suffixes (e.g. `_FAM_4` vs `_FAM_1`) denote genuinely different, unrelated HMM families rather than variants of the same model, so they should **not** be collapsed together.
:::

## 5. Heatmaps

### 5a. HMM counts per isolate — ggplot2

A tile heatmap of every isolate x HMM combination, coloured by how many copies of that HMM were found. Missing combinations are filled with 0 (absence) via `complete()`, so every isolate x HMM cell is represented even if it never appeared in the raw data.

```{r}
#| label: hmm-heatmap-ggplot
#| fig-width: 10
#| fig-height: 18

hmm_counts_long <- df %>%
  count(sample, hmm) %>%
  complete(sample, hmm, fill = list(n = 0))

ggplot(hmm_counts_long, aes(x = hmm, y = sample, fill = n)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_gradient(low = "grey95", high = "darkred", name = "Copies") +
  labs(x = "HMM model", y = "Isolate",
       title = "HMM copy number per isolate") +
  theme_minimal(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 4),   # 139 isolates - keep labels tiny
    panel.grid = element_blank()
  )
```

With 139 isolates the y-axis labels will be very small — this view is mainly for spotting overall patterns/outliers rather than reading individual isolate names off the plot. Use the `pheatmap` version below to reorder isolates by similarity, which usually makes patterns much easier to see, or filter `hmm_counts_long` to a subset of isolates first if you want a readable static figure.

### 5b. HMM presence/absence with clustering — pheatmap

`pheatmap` automatically performs hierarchical clustering on both rows (isolates) and columns (HMMs), so isolates with similar RM system profiles group together, and reveals which HMMs tend to co-occur. This uses presence/absence (0/1) rather than raw counts, which is usually cleaner for clustering — switch `binary_matrix` for `count_matrix` (commented below) if you want clustering based on copy number instead.

```{r}
#| label: hmm-heatmap-pheatmap
#| fig-width: 10
#| fig-height: 20

library(pheatmap)

# Wide matrix: isolates as rows, HMMs as columns
hmm_wide <- df %>%
  count(sample, hmm) %>%
  pivot_wider(names_from = hmm, values_from = n, values_fill = 0)

count_matrix <- hmm_wide %>%
  select(-sample) %>%
  as.matrix()
rownames(count_matrix) <- hmm_wide$sample

binary_matrix <- (count_matrix > 0) * 1   # presence/absence

pheatmap(
  binary_matrix,                 # swap for count_matrix to cluster on copy number
  color = c("grey95", "darkred"),
  legend_breaks = c(0, 1),
  legend_labels = c("Absent", "Present"),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize_row = 3,               # small - 139 isolates
  fontsize_col = 8,
  main = "HMM presence/absence across isolates (clustered)"
)
```

If `pheatmap` isn't installed, run `install.packages("pheatmap")` first.

## 6. HMM heatmap with metadata colour strips

This adds isolate-level metadata (e.g. species, source, collection site) as colour strips alongside the HMM heatmap from 5a. Uses `patchwork` to stack the strips and the heatmap so isolate rows line up exactly.

**Adjust the placeholders marked below to match your actual metadata file and column names.**

```{r}
#| label: hmm-heatmap-metadata
#| fig-width: 11
#| fig-height: 18

library(patchwork)

# ---- 1. Read metadata -----------------------------------------------
metadata_path <- "data/isolate_metadata.csv"   # <-- change to your actual metadata file path
metadata <- read_csv("~/Objective_3/Objective_3_RM_analysis/data/metadata_Apr26.csv", show_col_types = FALSE) %>%
  clean_names()   # -> isolate_id, farm_id, sample_site, total_length, total_contigs,
                   #    gc_percent, n50, l90, sbi, lactose, sorbitol, surcrose, stx_gene

# Join key: df$sample and metadata's isolate_id now match directly
# (both e.g. "EcCa437a") since the "fix_" prefix was stripped in Setup.
metadata <- metadata %>%
  rename(sample = isolate_id) %>%
  mutate(farm_id = as.factor(farm_id))   # farm ID is categorical, not numeric

# Columns to show as colour strips - add/remove as needed
annotation_vars <- c("sample_site", "sbi", "stx_gene")

# ---- 2. Join metadata onto the HMM long-format counts ----------------
hmm_counts_long_annot <- df %>%
  count(sample, hmm) %>%
  complete(sample, hmm, fill = list(n = 0)) %>%
  left_join(metadata, by = "sample")

# Fix isolate ordering so every panel below shares the same y-axis order
isolate_order <- sort(unique(hmm_counts_long_annot$sample))
hmm_counts_long_annot <- hmm_counts_long_annot %>%
  mutate(sample = factor(sample, levels = isolate_order))

# ---- 3. Main HMM heatmap ---------------------------------------------
main_heatmap <- ggplot(hmm_counts_long_annot, aes(x = hmm, y = sample, fill = n)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_gradient(low = "grey95", high = "darkred", name = "Copies") +
  labs(x = "HMM model", y = NULL) +
  theme_minimal(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 4),
    panel.grid = element_blank(),
    legend.position = "right"
  )

# ---- 4. One narrow colour-strip panel per annotation variable ---------
make_strip <- function(var) {
  strip_data <- hmm_counts_long_annot %>%
    distinct(sample, .data[[var]])

  ggplot(strip_data, aes(x = 1, y = sample, fill = .data[[var]])) +
    geom_tile(color = "white", linewidth = 0.1) +
    labs(x = NULL, y = NULL, fill = var) +
    theme_minimal(base_size = 9) +
    theme(
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      legend.position = "right"
    )
}

strip_plots <- lapply(annotation_vars, make_strip)

# ---- 5. Combine strips + main heatmap side by side --------------------
# Strips are narrow, main heatmap is wide - widths reflect that
combined_plot <- wrap_plots(
  c(strip_plots, list(main_heatmap)),
  nrow = 1,
  widths = c(rep(0.3, length(strip_plots)), 10)
)

combined_plot
```

If `patchwork` isn't installed, run `install.packages("patchwork")` first. The `make_strip()` helper works for any number of annotation variables — just add or remove names from `annotation_vars`, no other code changes needed.

## 7. System type heatmap

Same treatment as the HMM heatmap in Section 5, but using the `system_type_totals` table (RM_Type_I, RM_Type_II, RM_Type_IIG, Type_IV counts per isolate) instead of individual HMMs. With only 4-5 system type columns this is a much narrower heatmap than the HMM one.

### 7a. System type counts per isolate — ggplot2

```{r}
#| label: system-type-heatmap-ggplot
#| fig-width: 6
#| fig-height: 18

system_type_long <- system_type_totals %>%
  select(-Total) %>%
  pivot_longer(-sample, names_to = "system_type", values_to = "n")

ggplot(system_type_long, aes(x = system_type, y = sample, fill = n)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_gradient(low = "grey95", high = "darkblue", name = "Count") +
  labs(x = "System type", y = "Isolate",
       title = "RM system type counts per isolate") +
  theme_minimal(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 4),   # 139 isolates - keep labels tiny
    panel.grid = element_blank()
  )
```

### 7b. System type counts with clustering — pheatmap

```{r}
#| label: system-type-heatmap-pheatmap
#| fig-width: 8
#| fig-height: 20

system_type_matrix <- system_type_totals %>%
  select(-Total) %>%
  column_to_rownames("sample") %>%
  as.matrix()

pheatmap(
  system_type_matrix,
  color = colorRampPalette(c("grey95", "darkblue"))(50),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize_row = 3,               # small - 139 isolates
  fontsize_col = 9,
  main = "RM system type counts across isolates (clustered)"
)
```

## 8. Gene type heatmap (by isolate, not contig)

Same idea as Section 1's `gene_type_by_contig`, but collapsed across contigs so there's one row per isolate rather than one row per isolate x contig. This gives total counts of each gene type (RE, MT, SP, IIG) per isolate, regardless of which contig they're on.

```{r}
#| label: gene-type-by-isolate

gene_type_by_isolate <- df %>%
  count(sample, gene_type) %>%
  pivot_wider(names_from = gene_type, values_from = n, values_fill = 0) %>%
  arrange(sample)

gene_type_by_isolate
```

### 8a. Gene type counts per isolate — ggplot2

```{r}
#| label: gene-type-heatmap-ggplot
#| fig-width: 6
#| fig-height: 18

gene_type_long <- gene_type_by_isolate %>%
  pivot_longer(-sample, names_to = "gene_type", values_to = "n")

ggplot(gene_type_long, aes(x = gene_type, y = sample, fill = n)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_gradient(low = "grey95", high = "darkgreen", name = "Count") +
  labs(x = "Gene type", y = "Isolate",
       title = "Gene type counts per isolate") +
  theme_minimal(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 4),   # 139 isolates - keep labels tiny
    panel.grid = element_blank()
  )
```

### 8b. Gene type counts with clustering — pheatmap

```{r}
#| label: gene-type-heatmap-pheatmap
#| fig-width: 8
#| fig-height: 20

gene_type_matrix <- gene_type_by_isolate %>%
  column_to_rownames("sample") %>%
  as.matrix()

pheatmap(
  gene_type_matrix,
  color = colorRampPalette(c("grey95", "darkgreen"))(50),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize_row = 3,               # small - 139 isolates
  fontsize_col = 9,
  main = "Gene type counts across isolates (clustered)"
)
```