---
title: "Objective_3_RM_analysis"
format: html
editor: visual
---

## Setup

Adjust `input_path` below to point at your combined RM annotation CSV.

```{r}
#| label: libraries

library(tidyverse)
library(janitor)
library(gt)

```


```{r}
#| label: setup

input_path <- "~/Objective_3/Objective_3_RM_analysis/data/all.rm.genes.concatRpt.tsv"   # <-- change this to your actual file path

df <- read_tsv(input_path, show_col_types = FALSE) %>%
  clean_names()   # -> sample, operon, gene, system_type, gene_type,
                  #    hmm, evalue, predicted_methylation,
                  #    prediction_confidence, rebase_homolog,
                  #    homolog_identity_percent, homolog_methylation,
                  #    homolog_motif

# derive contig from the Gene locus tag, e.g. "contig_1_2542" -> "contig_1"
df <- df %>%
  mutate(contig = str_extract(gene, "^contig_[0-9]+"))

n_isolates <- n_distinct(df$sample)
n_hmm <- n_distinct(df$hmm)
```


This dataset covers **`r n_isolates`** isolates and **`r n_hmm`** distinct HMM models.

## 1. Gene type counts per isolate, per contig

Counts of each gene type (RE, MT, SP, IIG) found on each contig, within each isolate.

```{r}
#| label: gene-type-by-contig

gene_type_by_contig <- df %>%
  count(sample, contig, gene_type) %>%
  pivot_wider(names_from = gene_type, values_from = n, values_fill = 0) %>%
  arrange(sample, contig)

gene_type_by_contig
```


## 2. System type totals per isolate

Counts of each system type (RM_Type_I, RM_Type_II, RM_Type_IIG, Type_IV) per isolate, with a row total.

```{r}
#| label: system-type-totals

system_type_totals <- df %>%
  count(sample, system_type) %>%
  pivot_wider(names_from = system_type, values_from = n, values_fill = 0) %>%
  mutate(Total = rowSums(across(where(is.numeric)))) %>%
  arrange(sample)

system_type_totals
```


## 3. Operon-level summary

One row per isolate x operon, showing the operon's system type(s), contig(s), gene types present, gene count, and whether any gene in the operon has a methylation call.

```{r}
#| label: operon-summary

operon_summary <- df %>%
  group_by(sample, operon) %>%
  summarise(
    contigs        = paste(sort(unique(contig)), collapse = ";"),
    operon_type    = paste(sort(unique(system_type)), collapse = ";"),
    gene_types     = paste(sort(unique(gene_type)), collapse = ";"),
    n_genes        = n(),
    has_methylation_call = any(!is.na(predicted_methylation) & predicted_methylation != ""),
    .groups = "drop"
  ) %>%
  arrange(sample, operon)

operon_summary
```

## 4. HMM inventory

### 4a. Overall counts across all isolates

```{r}
#| label: hmm-overall

hmm_inventory_overall <- df %>%
  count(hmm, sort = TRUE) %>%
  rename(n_occurrences = n)

hmm_inventory_overall %>%
  gt() %>%
  tab_header(title = "Distinct HMM models identified across all isolates")
```

### 4b. Per-isolate breakdown

```{r}
#| label: hmm-by-isolate

hmm_inventory_by_isolate <- df %>%
  count(sample, hmm) %>%
  pivot_wider(names_from = hmm, values_from = n, values_fill = 0) %>%
  arrange(sample)

hmm_inventory_by_isolate
```

::: {.callout-note}
The raw `hmm` column already condenses to ~16 distinct models across the whole dataset — no further grouping needed. Note that HMM family suffixes (e.g. `_FAM_4` vs `_FAM_1`) denote genuinely different, unrelated HMM families rather than variants of the same model, so they should **not** be collapsed together.
:::

## Export tables

```{r}
#| label: export

write_csv(gene_type_by_contig,      "data/1_gene_type_by_contig.csv")
write_csv(system_type_totals,       "data/2_system_type_totals.csv")
write_csv(operon_summary,           "data/3_operon_summary.csv")
write_csv(hmm_inventory_overall,    "data/4a_hmm_inventory_overall.csv")
write_csv(hmm_inventory_by_isolate, "data/4b_hmm_inventory_by_isolate.csv")
```
